---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CFtime

<!-- badges: start -->
<!-- badges: end -->

CFtime is an R package that supports working with [CF Metadata Conventions](http://cfconventions.org)
time coordinates, specifically geared to time-referencing data sets 
of climate change projections such as those produced under the [World Climate
Research Programme](https://www.wcrp-climate.org) and re-analysis data such as
ERA5.

The data sets include in their metadata a datum, a point in time from which 
other points in time are calculated. This datum takes the form of `days since
1949-12-01`, with each data source (CMIP generation, model, etc) having its own
datum. The data itself has a dimension attribute of "time" with offset values 
such as 43289. To convert this "offset" to a date, using a specific calendar, is
what this package does. Given that most calendars supported by the CF Metadata
Conventions are not compatible with `POSIXt`, this conversion is not trivial.
That it is important account for these differences is easily demonstrated:

``` {r, example_1}
library(CFtime)

# POSIXt calculations on a standard calendar
as.Date("1949-12-01") + 43289

# CFtime calculation on a "360_day" calendar
CFtimestamp(CFts(CFdatum("days since 1949-12-01", "360_day"), 43289))
```

That's a difference of nearly 15 months! (And yes, 30 February is a valid date
on a `360_day` calendar.)

All defined calendars of the CF Metadata Convention are supported:

 - `standard` or `gregorian`
 - `proleptic_gregorian`
 - `julian`
 - `365_day` or `no_leap`
 - `366_day` or `all_leap`
 - `360_day`

Use of custom calendars is not supported. This package is also not suitable for
paleo-calendars. Time periods prior to the introduction of the Gregorian calendar
(1582-10-15) may be used but there are no special provisions for it. Finally,
there is no specific consideration for the year 0 (which does not exist in any of
the above calendars). Given that climate change projections are typically made
from 1850-01-01 onwards, these limitations should not be of any practical 
concern.

This package IS NOT intended to support the full date and time functionality of
the CF Metadata Conventions. Instead, it facilitates use of a suite of models of
climate change projections that use different calendars in a consistent manner.

This package is particularly useful for working with climate change projections
having a daily or higher resolution, but it will work equally well on data with
a lower resolution.

Timestamps are generated using the [ISO8601 standard](https://en.wikipedia.org/wiki/ISO_8601).

Calendar-aware factors can be generated to support processing of data using `tapply()`
and similar functions.

## Installation

You can install the development version of CFtime from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pvanlaake/CFtime")
```

The package has not yet been uploaded to CRAN.

## Basic operation

The package contains two classes, `CFdatum` to describe the time coordinate
system, including its calendar; and `CFts` to hold the time coordinate values
that are offset from the `CFdatum` to represent instants in time. These
two classes operate on the data in the file of interest, here a CORDEX file of
precipitation for the Central America domain:

```{r example_2, eval = FALSE}
library(CFtime)
library(ncdf4)
nc <- nc_open("~/CC/CORDEX/CAM-22/RCP2.6/pr_CAM-22_MOHC-HadGEM2-ES_rcp26_r1i1p1_GERICS-REMO2015_v1_day_20060101-20101230.nc")
datum <- CFdatum(nc$dim$time$units, nc$dim$time$calendar)
time <- CFts(datum, nc$dim$time$vals)
time
nc_close(nc)
```

Note that the information of interest (`nc$dim$time$units`, etc) is read out of
the file "blindly", without checking for available dimensions or attributes.
This can be done because the `time` dimension and its attributes `units` and
`calendar` are required by the CF Metadata Conventions. Should this fail, then
your data set is not compliant. You could still use this package if the required
information is contained in your file but using a different dimension name or
different attribute names.

## Typical workflow

In a typical process, you would combine multiple data files into a single data
set to analyze a feature of interest. To continue the previous example of
precipitation in the Central America domain using CORDEX data, you can calculate
the precipitation per month for the period 2041 - 2050 as follows:

```{r example_3, eval = FALSE}
library(CFtime)
library(ncdf4)
library(abind)
# Open the files - one would typically do this in a loop
nc2041 <- nc_open("~/CC/CORDEX/CAM-22/RCP2.6/pr_CAM-22_MOHC-HadGEM2-ES_rcp26_r1i1p1_GERICS-REMO2015_v1_day_20410101-20451230.nc")
nc2046 <- nc_open("~/CC/CORDEX/CAM-22/RCP2.6/pr_CAM-22_MOHC-HadGEM2-ES_rcp26_r1i1p1_GERICS-REMO2015_v1_day_20460101-20501230.nc")

# Create the datum from the first file
# All files have an identical datum as per the CORDEX specifications
datum <- CFdatum(nc2041$dim$time$units, nc2041$dim$time$calendar)

# Create the time series object from the first file, then add the time values from 
# the remaining files
time <- CFts(datum, nc2041$dim$time$vals) + as.vector(nc2046$dim$time$vals)

# Grab the data from the files and merge the arrays into one
pr <- abind(ncvar_get(nc2041, "pr"), ncvar_get(nc2046, "pr"))
nc_close(nc2041)
nc_close(nc2046)

# Optionally - Set the time dimension to the timestamps from the time object
dimnames(pr)[[3]] <- CFtimestamp(time)

# Create the month factor from the time object
f_month <- CFfactor(time, "month")

# Now sum the daily data to monthly data
# Dimensions 1 and 2 are longitude and latitude, the third dimension is time
pr_month <- aperm(apply(pr, 1:2, tapply, f_month, sum), c(2, 3, 1))
dimnames(pr_month)[[3]] <- levels(f_month)
```

## Coverage

This package has been tested with the following data sets:

 - ERA5 (including multiple variables, levels, and mixed ERA5/ERA5T data)
 - CMIP5
 - CORDEX
 - CMIP6

The package also operates on geographical and/or temporal subsets of data sets
so long as the subsetted data complies with the CF Metadata Conventions. This
includes subsetting in the [Climate Data Store](https://cds.climate.copernicus.eu/#!/home).
Subsetted data from [Climate4Impact](https://climate4impact.eu/impactportal/general/index.jsp)
is not automatically supported because the dimension names are not compliant
with the CF Metadata Conventions, use the corresponding dimension names instead.
